---
title: "NBA ggmemo Demo — 2025-26 Season"
subtitle: "Annotated charts built with ggplot2 + ggmemo"
format:
  html:
    theme: cosmo
    toc: true
    code-fold: true
    fig-width: 10
    fig-height: 6
    fig-dpi: 200
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(ggmemo)

nba <- read.csv("nba_top30_2025_26.csv")
```

## Top 10 Scorers

Horizontal bar chart with `annotate_callout()` highlighting Luka Dončić as scoring champ and Nikola Jokić's triple-double averages.

```{r}
#| label: fig-scorers
#| fig-cap: "Points per game, 2025-26 regular season"
#| fig-height: 6

top10 <- nba[1:10, ]
top10$PLAYER <- factor(top10$PLAYER, levels = rev(top10$PLAYER))

small_arrow <- grid::arrow(length = grid::unit(0.12, "inches"))

ggplot(top10, aes(y = PLAYER, x = PTS)) +
  geom_col(fill = "#1D428A", width = 0.7) +
  annotate_callout(
    top10,
    where = RANK == 1,
    label = "Scoring champ",
    position = "top-right",
    nudge = c(2, 0.4),
    arrow = small_arrow
  ) +
  annotate_callout(
    top10,
    where = PLAYER == "Nikola Jokić",
    label = "Triple-double machine\n(12.9 REB, 10.7 AST)",
    position = "top-right",
    nudge = c(3, 0.4),
    size = 3.5,
    arrow = small_arrow
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_cartesian(clip = "off") +
  labs(y = NULL, x = "PPG") +
  theme_minimal(base_size = 14) +
  theme(plot.margin = margin(10, 80, 10, 10))
```

## Scoring vs Playmaking

Bubble scatter plot (size = rebounds) with four `annotate_callout()` labels picking out Luka, Jokić, rookie Cooper Flagg, and Wemby's 3.1 blocks per game.

```{r}
#| label: fig-scatter
#| fig-cap: "Points vs assists per game — bubble size is rebounds"
#| fig-height: 7

small_arrow <- grid::arrow(length = grid::unit(0.12, "inches"))

ggplot(nba, aes(x = AST, y = PTS)) +
  geom_point(aes(size = REB), alpha = 0.7, color = "#1D428A") +
  annotate_callout(
    nba,
    where = PLAYER == "Nikola Jokić",
    label = "Jokić",
    position = "top-left",
    nudge = c(0.8, 1.5),
    arrow = small_arrow
  ) +
  annotate_callout(
    nba,
    where = PLAYER == "Luka Dončić",
    label = "Luka",
    position = "top-left",
    nudge = c(0.8, 1.5),
    arrow = small_arrow
  ) +
  annotate_callout(
    nba,
    where = PLAYER == "Cooper Flagg",
    label = "Flagg (rookie)",
    position = "bottom-left",
    nudge = c(1.0, 2.0),
    arrow = small_arrow
  ) +
  annotate_callout(
    nba,
    where = PLAYER == "Victor Wembanyama",
    label = "Wemby\n(3.1 BLK)",
    position = "top-right",
    nudge = c(0.6, 1.5),
    size = 3.5,
    arrow = small_arrow
  ) +
  scale_size_continuous(range = c(2, 8), name = "Rebounds") +
  labs(x = "Assists per game", y = "Points per game") +
  theme_minimal(base_size = 14)
```

## Field Goal Efficiency

`annotate_change(format = "points")` shows the 11.1 percentage-point gap between Zion Williamson (60.0%) and Jalen Johnson (48.9%).

```{r}
#| label: fig-fg-pct
#| fig-cap: "Top 10 field goal percentage, 2025-26 regular season"

fg_top <- nba[order(-nba$FG_PCT), ][1:10, ]
fg_top$FG_PCT_100 <- fg_top$FG_PCT * 100
fg_top$PLAYER <- factor(fg_top$PLAYER, levels = fg_top$PLAYER)

fg_simple <- fg_top[, c("PLAYER", "FG_PCT_100")]

ggplot(fg_top, aes(x = PLAYER, y = FG_PCT_100)) +
  geom_col(fill = "#C8102E", width = 0.6) +
  annotate_change(
    fg_simple,
    from = PLAYER == levels(fg_simple$PLAYER)[1],
    to = PLAYER == levels(fg_simple$PLAYER)[10],
    value = FG_PCT_100,
    format = "points"
  ) +
  labs(x = NULL, y = "FG%") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Efficiency Rating

Lollipop (Cleveland dot) chart showing `annotate_callout()` works beyond bar charts. SGA is called out as 2025-26 MVP, while Luka's callout uses `colour = "#C8102E"` to highlight the efficiency gap despite leading in scoring.

```{r}
#| label: fig-efficiency
#| fig-cap: "Efficiency rating, 2025-26 regular season"
#| fig-height: 7

nba_eff <- nba[order(-nba$EFF), ]
nba_eff$PLAYER <- factor(nba_eff$PLAYER, levels = rev(nba_eff$PLAYER))

ggplot(nba_eff[1:15, ], aes(x = EFF, y = PLAYER)) +
  geom_segment(aes(x = 0, xend = EFF, yend = PLAYER),
               color = "grey70", linewidth = 0.4) +
  geom_point(size = 4, color = "#552583") +
  annotate_callout(
    nba_eff[1:15, ],
    where = PLAYER == "Shai Gilgeous-Alexander",
    label = "2025-26 MVP",
    position = "bottom-right",
    nudge = c(3, 0.8),
    size = 3.5
  ) +
  annotate_callout(
    nba_eff[1:15, ],
    where = PLAYER == "Luka Dončić",
    label = "Scoring champ,\nlower efficiency",
    position = "top-right",
    nudge = c(3, 0.8),
    size = 3.5,
    colour = "#C8102E"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "NBA Efficiency Rating — Top 15 Players",
    subtitle = "Points + rebounds + assists + steals + blocks - missed shots - turnovers",
    x = "Efficiency Rating", y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        plot.margin = margin(10, 60, 10, 10))
```

## Three-Point Shooting

Bubble scatter of 3-point attempts vs accuracy, with three `annotate_callout()` labels highlighting the volume leader, the best volume-accuracy combo (Murray), and Flagg's limited range in red.

```{r}
#| label: fig-threes
#| fig-cap: "3-point attempts vs accuracy — bubble size is PPG"
#| fig-height: 7

small_arrow <- grid::arrow(length = grid::unit(0.12, "inches"))

ggplot(nba, aes(x = FG3A, y = FG3_PCT * 100)) +
  geom_point(aes(size = PTS), alpha = 0.6, color = "#006BB6") +
  annotate_callout(
    nba,
    where = FG3A == max(nba$FG3A),
    label = "Most 3PA/game",
    position = "bottom-left",
    nudge = c(0.5, 1.5),
    arrow = small_arrow
  ) +
  annotate_callout(
    nba,
    where = PLAYER == "Jamal Murray",
    label = "Murray: best combo\n(7.5 3PA, 43.5%)",
    position = "top-left",
    nudge = c(0.8, 1.5),
    size = 3.5,
    arrow = small_arrow
  ) +
  annotate_callout(
    nba,
    where = PLAYER == "Cooper Flagg",
    label = "Flagg: limited range\n(3.5 3PA, 29.5%)",
    position = "bottom-right",
    nudge = c(0.5, 2),
    size = 3.5,
    colour = "#C8102E",
    arrow = small_arrow
  ) +
  scale_size_continuous(range = c(2, 7), name = "PPG") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(
    title = "Three-Point Shooting: Volume vs Accuracy",
    subtitle = "Bubble size = points per game",
    x = "3-Point Attempts per Game",
    y = "3-Point %"
  ) +
  theme_minimal(base_size = 14)
```

## Top Playmakers

`annotate_change(format = "absolute")` shows the raw assist gap between Jokić and SGA, while `annotate_callout()` highlights Cade Cunningham's breakout playmaking season.

```{r}
#| label: fig-assists
#| fig-cap: "Top 10 assists per game, 2025-26 regular season"

ast_top <- nba[order(-nba$AST), ][1:10, ]
ast_top$PLAYER <- factor(ast_top$PLAYER, levels = ast_top$PLAYER)
ast_simple <- ast_top[, c("PLAYER", "AST")]

ggplot(ast_top, aes(x = PLAYER, y = AST)) +
  geom_col(fill = "#007A33", width = 0.6) +
  annotate_change(
    ast_simple,
    from = PLAYER == levels(ast_simple$PLAYER)[1],
    to = PLAYER == levels(ast_simple$PLAYER)[10],
    value = AST,
    format = "absolute"
  ) +
  annotate_callout(
    ast_top,
    where = PLAYER == "Cade Cunningham",
    label = "Most improved\nplaymaker",
    position = "bottom-right",
    nudge = c(0.5, 0.5),
    size = 3.5
  ) +
  labs(
    title = "Top 10 Playmakers — Assists per Game",
    subtitle = "2025-26 regular season",
    x = NULL, y = "APG"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Rebounding: Wemby vs Flagg

Head-to-head comparison using `annotate_change(format = "percent")` for the percentage difference, combined with `annotate_callout()` on both bars.

```{r}
#| label: fig-wemby-flagg
#| fig-cap: "Per-game rebounds, Wembanyama vs Flagg"

duo <- nba[nba$PLAYER %in% c("Victor Wembanyama", "Cooper Flagg"), ]
duo$PLAYER <- factor(duo$PLAYER, levels = c("Cooper Flagg", "Victor Wembanyama"))
duo_simple <- duo[, c("PLAYER", "REB")]

ggplot(duo, aes(x = PLAYER, y = REB)) +
  geom_col(fill = c("#00538C", "#C4CED4"), width = 0.5) +
  annotate_change(
    duo_simple,
    from = PLAYER == "Cooper Flagg",
    to = PLAYER == "Victor Wembanyama",
    value = REB,
    format = "percent"
  ) +
  annotate_callout(
    duo,
    where = PLAYER == "Victor Wembanyama",
    label = "Elite rebounder at 7'4\"",
    position = "top-left",
    nudge = c(0.15, 0.6)
  ) +
  annotate_callout(
    duo,
    where = PLAYER == "Cooper Flagg",
    label = "Rookie: 6.7 RPG",
    position = "top-left",
    nudge = c(0.15, 0.6),
    size = 3.5
  ) +
  labs(
    title = "Rebounding: Wemby vs Flagg",
    subtitle = "Per-game rebounds, 2025-26 season",
    x = NULL, y = "Rebounds per Game"
  ) +
  theme_minimal(base_size = 14)
```

::: {.callout-note appearance="minimal"}
Data from [stats.nba.com](https://www.nba.com/stats) — 2025-26 regular season, per-game averages.
:::
